---
title: 'Input/output'
author: "Kenneth B. Hoehn"
date: '`r Sys.Date()`'
output:
  pdf_document:
    dev: pdf
    fig_height: 4
    fig_width: 7.5
    highlight: pygments
    toc: yes
  html_document:
    fig_height: 4
    fig_width: 7.5
    highlight: pygments
    theme: readable
    toc: yes
  md_document:
    fig_height: 4
    fig_width: 7.5
    preserve_yaml: no
    toc: yes
geometry: margin=1in
fontsize: 11pt
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Input and Output}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

This vignette covers how data gets into Dowser and the different ways Dowser
objects and trees can be saved. These formats primarily follow the [Adaptive Immune Receptor Repertoire (AIRR) Community standard formats](https://docs.airr-community.org).
Specifically this will cover AIRR TSV input, as well as AIRR Trees and Clones JSON format
(`writeTreesJSON`/`readTreesJSON`), plain R serialization (`saveRDS`/`readRDS`),
Newick tree export (`exportTrees`), and FASTA sequence export/import
(`writeFasta`/`readFasta`).

## Reading AIRR Rearrangement sequencing data files

The best input format for Dowser is the [AIRR Rearrangement](https://docs.airr-community.org/en/latest/datarep/rearrangements.html)
TSV format. The `airr` package's `read_rearrangement` function reads such a
file into a data frame:

```{r, eval=FALSE, warning=FALSE, message=FALSE}
library(airr)
library(dowser)

# Read in an AIRR-formatted TSV file of Ig/TCR rearrangements
airr_data <- read_rearrangement("sequences.tsv")
```

To keep this vignette self-contained, we'll instead use the `ExampleAirr`
data object included with Dowser, which is the kind of data frame
`read_rearrangement` would return.

```{r, eval=TRUE, warning=FALSE, message=FALSE}
library(airr)
library(dowser)

# load example AIRR data, as if read in with read_rearrangement
data(ExampleAirr)

# subset data for this example
ExampleAirr <- ExampleAirr[ExampleAirr$clone_id %in% c("3170", "3184"),]
```

From here, `formatClones` groups sequences into clones and reconstructs
germlines, and `getTrees` builds a lineage tree for each clone (see the
[Build Lineage Trees vignette](Building-Trees-Vignette.html) for details on
both steps). 

Before this step, it's important that the BCR sequences are already grouped into
clones beforehand (they will have a `clone_id` column). This can be accomplished 
using [SCOPer](https://scoper.readthedocs.io). Further, it is imporant that each 
clonal germline V/J has has been reconstructed, which can be accomplished using 
(createGermlines)[https://dowser.readthedocs.io/en/stable/vignettes/Germlines-Vignette/].
This will generate a `germline_alignment_d_mask` column. To see all of these steps 
together, check out one of the full [Immcantation tutorials](https://immcantation.readthedocs.io/en/stable/getting_started/getting-started.html).

```{r, eval=TRUE, warning=FALSE, message=FALSE}
# Process example data into proper format, store isotype (optional)
clones <- formatClones(ExampleAirr, traits="c_call")

# Build maximum parsimony trees for each clone
trees <- getTrees(clones, nproc=1)

print(trees)
```

`trees` is a standard Dowser object: a tibble with one row per clone with an
`airrClone` object in the `data` column and a tree (`ape::phylo`, or
`treeio::treedata` for time trees) in the `trees` column. The rest of this
vignette covers ways to save and reload this object.

## Saving and loading Dowser objects in AIRR Clone JSON format

To save an entire Dowser object, including trees, use `writeTreesJSON` to write 
a Dowser object to an [AIRR Clone](https://docs.airr-community.org/en/latest/datarep/clone.html)
JSON file. Conversely, `readTreesJSON` reads it back. This works for trees built with
any `getTrees` `build` option, as well as time trees from `getTimeTrees`/
`getTimeTreesIterate`. By default, `writeTreesJSON` reloads the file it just
wrote and checks that it matches the original object before returning.

```{r, eval=TRUE, warning=FALSE, message=FALSE}
# Write tree object to a JSON file, checking it reads back identically
writeTreesJSON(trees, "trees.json")

# Read it back in
trees_json <- readTreesJSON("trees.json")

print(trees_json)
```

Because it's plain JSON with a documented schema, this format is useful for
archiving trees long-term, sharing them with collaborators who aren't using
R, or loading them into other tools.

**Note** that internal node labels are not preserved when inputting/outputting in this
schema, but this is only relevant if you're manually selecting particular nodes,
such as [reconstructing internal node sequences](https://dowser.readthedocs.io/en/stable/vignettes/Sequences-Vignette/).
All other aspects of the Dowser object are preserved and checked on output.

## Saving and loading Dowser objects as RDS

If you're staying within R, `saveRDS`/`readRDS` is the simplest way to save 
a Dowser object exactly as-is, with no format translation, although it is a binary
file that is unreadable outside of R.

```{r, eval=TRUE, warning=FALSE, message=FALSE}
# Save tree object to an RDS file
saveRDS(trees, "trees.rds")

# Read it back in
trees_rds <- readRDS("trees.rds")

print(trees_rds)
```

## Exporting trees in Newick format

`exportTrees` writes the trees in a Dowser object in order to a single Newick file,
one tree per line, using `ape::write.tree`.

Alternatively, the list of tree objects in the `trees` column can be exported using
many other output functions, for example `ape::write.tree`.

```{r, eval=TRUE, warning=FALSE, message=FALSE}
# Export trees to a Newick file
exportTrees(trees, "trees.newick")

# write a single tree
ape::write.tree(trees$trees[[1]], "clone1.tree")
```

## Exporting and reading sequences as FASTA

`writeFasta` writes a named list of sequences to a FASTA file, and
`readFasta` reads a FASTA file back into a named list. These are generic
sequence I/O functions and provided for convenience.

```{r, eval=TRUE, warning=FALSE, message=FALSE}
# Get sequences and sequence IDs for the first clone
seqs <- clones$data[[1]]@data$sequence
names(seqs) <- clones$data[[1]]@data$sequence_id

# Write them to a FASTA file
writeFasta(seqs, "clone_3170.fasta")

# Read them back in
seqs_read <- readFasta("clone_3170.fasta")

print(names(seqs_read))
```

Alternatively, `dfToFasta` writes sequences straight from a data frame or tibble to a
FASTA file, without needing to build a named list first. By default it
reads sequence IDs and sequences from the `sequence_id`/`sequence`
columns, strips IMGT gap characters (`.`), and can append extra columns
to each sequence's header.

```{r, eval=TRUE, warning=FALSE, message=FALSE}
# Get the data frame of sequences for the first clone
df <- clones$data[[1]]@data

# Write to FASTA, tagging each header with its isotype
dfToFasta(df, "clone_3170_df.fasta", columns="c_call")

cat(readLines("clone_3170_df.fasta")[1:2], sep="\n")
```

Use `id`/`seq` to point at differently-named columns, and `imgt_gaps=TRUE`
to keep IMGT gaps rather than stripping them.

## Making and saving tree plots

Now that trees are built and saved, see the
[Plotting Trees vignette](https://dowser.readthedocs.io/en/stable/vignettes/Plotting-Trees-Vignette/) 
for how to visualize trees and save tree plots as files.
